---
title: 'DS 202 - Project Proposal'
author: "Save Gaza"
output: html_document
date: "03-29-2024"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of agricultural economics statistics in the US

We want to analyze the agricultural economics statistics in the US. We are using [real world data](https://www.nass.usda.gov/datasets/) from the USDA NASS. 
Our specific dataset is `qs.economics_20240329.gz`

## Team Members

Hazer Becic, Ashraf Shaikh Mohammed, Luis Hinkhouse, Mohamed Bashier, Mazin Bashier.

## Data

Our data is present as a zipped txt file in tab separated format. It appears the data is relatively clean on the first glance, so we assume not much intensive data cleaning needs to be done.

There's [accompanying documentation](https://quickstats.nass.usda.gov/src/glossary.pdf) to understand certain values in our dataset.

```{r, message=FALSE, echo=FALSE}

library(rvest)
library(tidyverse)
library(ggplot2)

url <- "https://www.nass.usda.gov/datasets/"
html_content <- read_html(url)
text_strings <- html_content %>% html_nodes("div") %>% lapply(function(x) {x %>% html_text() %>% trimws()})
pattern <- "^qs\\.economics_\\d+\\.txt\\.gz$"
matches <- regmatches(text_strings, regexpr(pattern, text_strings))
print(matches)
data_url <- paste("https://www.nass.usda.gov/datasets/", matches, sep = "")
print(data_url)
```

```{r, message=FALSE, echo=FALSE}

options(timeout=120)

usda_dataset <- matches
#usda_dataset <- 'qs.economics_20240329.txt.gz'
if (!file.exists(usda_dataset))
{
  download.file(data_url, destfile = usda_dataset, mode = 'wb')
}

# This step takes a while so please be patient
suppressWarnings(data_usda <- read.delim(gzfile(usda_dataset)))

```

```{r, message=FALSE, echo=FALSE}
#nrow(data_usda)
# We have a whopping 11,695,604 rows in our dataset

#length(colnames(data_usda))
# We have 39 columns in our dataset

#dim(data_usda)

colnames(data_usda)
```


We see we have 39 columns containing lots of valuable information. We shall now see a preview of certain columns.

```{r, message=FALSE, echo=FALSE}
# Preview certain columns
data_usda_preview <- data_usda %>% 
  select(COMMODITY_DESC, YEAR, STATE_NAME, VALUE, UNIT_DESC, COUNTRY_NAME)

str(data_usda_preview)

data_usda_preview %>% head(n = 20)


```


```{r, message=FALSE, echo=FALSE}

# Select most important columns

data_usda <- data_usda %>% select(SOURCE_DESC, GROUP_DESC,
                     COMMODITY_DESC, STATISTICCAT_DESC, AGG_LEVEL_DESC, DOMAIN_DESC,
                     COUNTRY_NAME, STATE_NAME,
                     FREQ_DESC, YEAR, REFERENCE_PERIOD_DESC, VALUE, UNIT_DESC)

# REFERENCE_PERIOD_DESC - for months
# VALUE - remove D's, replace commas
# FREQ_DESC -
# YEAR -
# COMMODITY_DESC -
# STATISTICCAT_DESC - higher level of DOMAIN_DESC
# COUNTRY_NAME -
# STATE_NAME - ignore "" empty state; US TOTAL and OTHER STATES for statewise analysis
# UNIT_DESC -
# GROUP_DESC -
# SOURCE_DESC - let's only care about census data
# AGG_LEVEL_DESC -
# DOMAIN_DESC - maybe but mostly yes, seems more detailed than STATISTICCAT_DESC

# CLASS_DESC maybe, not so important
# PRODN_PRACTICE_DESC maybe, not so important
# UTIL_PRACTICE_DESC maybe, not so important



data_usda <- data_usda %>% filter(SOURCE_DESC == "CENSUS",
                                  COUNTRY_NAME == "UNITED STATES",
                                  VALUE != "(D)") %>% select(-c(SOURCE_DESC, COUNTRY_NAME))

data_usda <- data_usda %>% filter(STATE_NAME != "",
                                  REFERENCE_PERIOD_DESC == "YEAR") %>% select(-c(REFERENCE_PERIOD_DESC))

data_usda <- data_usda %>% select(-c(FREQ_DESC))


#head(data_usda$VALUE, 200)

```

```{r, message=FALSE, echo=FALSE}

colnames(data_usda)

#data_usda %>% select(-c(VALUE, STATISTICCAT_DESC, DOMAIN_DESC, COMMODITY_DESC)) %>% sapply(unique)

data_usda_test <- data_usda %>% filter(GROUP_DESC == "ENERGY"
#                                       ,COMMODITY_DESC == "AG LAND",
#                                       ,STATISTICCAT_DESC == "AREA",
#                                       ,DOMAIN_DESC == "AREA OPERATED" | DOMAIN_DESC == "AREA" | DOMAIN_DESC == "AREA HARVESTED"
)


#unique(data_usda$YEAR)
#unique(data_usda$UNIT_DESC)
print("------GROUP_DESC------")
unique(data_usda_test$GROUP_DESC)
print("------COMMODITY_DESC------")
unique(data_usda_test$COMMODITY_DESC)
print("------STATISTICCAT_DESC------")
unique(data_usda_test$STATISTICCAT_DESC)
print("------AGG_LEVEL_DESC------")
unique(data_usda_test$AGG_LEVEL_DESC)
print("------DOMAIN_DESC------")
unique(data_usda_test$DOMAIN_DESC)
print("------UNIT_DESC------")
unique(data_usda$UNIT_DESC)

#data_usda_test <- data_usda_test %>% filter(STATE_NAME == "IOWA", AGG_LEVEL_DESC == "STATE")
dim(data_usda_test)

# do like convert years, states to factors, etc.
# now our data is ready for analysis
```

## Questions to be addressed

We want to see what agricultural economic sectors are present in various states in the US as well as their growth/decline over the years.

```{r, message=FALSE, echo=FALSE}

data_usda_AG_LAND_AREA <- data_usda %>% filter(GROUP_DESC == "FARMS & LAND & ASSETS"
                                       ,COMMODITY_DESC == "AG LAND"
                                       ,STATISTICCAT_DESC == "AREA"
                                       )

data_usda_AG_LAND_AREA_IOWA <- data_usda_AG_LAND_AREA %>% filter(STATE_NAME == "IOWA", AGG_LEVEL_DESC == "STATE"
                                                                 ,UNIT_DESC == "ACRES"
                                                                 )

unique(data_usda_AG_LAND_AREA_IOWA$DOMAIN_DESC)

data_usda_AG_LAND_AREA_IOWA <- data_usda_AG_LAND_AREA_IOWA %>% filter(
  DOMAIN_DESC == "AREA CROPLAND, HARVESTED" | DOMAIN_DESC == "AREA OPERATED AND AREA CROPLAND, HARVESTED"
)


#View(data_usda_AG_LAND_AREA_IOWA)

# only do numeric conversion here
data_usda_AG_LAND_AREA_IOWA$VALUE <- as.numeric(gsub("[^0-9.]", "", data_usda_AG_LAND_AREA_IOWA$VALUE))

testt <- data_usda_AG_LAND_AREA_IOWA %>%
  group_by(YEAR, DOMAIN_DESC) %>%
  summarize(land_area = sum(VALUE)) 

#View(testt)

testt %>% ggplot(aes(x = factor(YEAR), y = land_area)) + geom_bar(stat = "identity", aes(fill = DOMAIN_DESC), position = "dodge") #+ coord_flip()


#data_usda_AG_LAND_AREA_IOWA %>% ggplot(aes(x = YEAR, y = VALUE, color = DOMAIN_DESC)) + geom_point()

```

```{r, message=FALSE, echo=FALSE}

data_usda_AG_LAND_AREA <- data_usda %>% filter(GROUP_DESC == "FARMS & LAND & ASSETS"
                                       ,COMMODITY_DESC == "AG LAND"
                                       ,STATISTICCAT_DESC == "AREA"
                                       )

data_usda_AG_LAND_AREA_NEBRASKA <- data_usda_AG_LAND_AREA %>% filter(STATE_NAME == "NEBRASKA", AGG_LEVEL_DESC == "STATE"
                                                                 ,UNIT_DESC == "ACRES"
                                                                 )

unique(data_usda_AG_LAND_AREA_NEBRASKA$DOMAIN_DESC)

data_usda_AG_LAND_AREA_NEBRASKA <- data_usda_AG_LAND_AREA_NEBRASKA %>% filter(
  DOMAIN_DESC == "AREA CROPLAND, HARVESTED" | DOMAIN_DESC == "AREA OPERATED AND AREA CROPLAND, HARVESTED"
)


#View(data_usda_AG_LAND_AREA_IOWA)

# only do numeric conversion here
data_usda_AG_LAND_AREA_NEBRASKA$VALUE <- as.numeric(gsub("[^0-9.]", "", data_usda_AG_LAND_AREA_NEBRASKA$VALUE))

testt <- data_usda_AG_LAND_AREA_NEBRASKA %>%
  group_by(YEAR, DOMAIN_DESC) %>%
  summarize(land_area = sum(VALUE)) 

#View(testt)

testt %>% ggplot(aes(x = factor(YEAR), y = land_area)) + geom_bar(stat = "identity", aes(fill = DOMAIN_DESC), position = "dodge") #+ coord_flip()


#data_usda_AG_LAND_AREA_IOWA %>% ggplot(aes(x = YEAR, y = VALUE, color = DOMAIN_DESC)) + geom_point()

```


```{r, message=FALSE, echo=FALSE}

data_usda_AG_LAND_AREA_2 <- data_usda %>% filter(GROUP_DESC == "FARMS & LAND & ASSETS"
                                       ,COMMODITY_DESC == "AG LAND"
                                       ,STATISTICCAT_DESC == "AREA"
                                       )

data_usda_AG_LAND_AREA_2 <- data_usda_AG_LAND_AREA_2 %>% filter(AGG_LEVEL_DESC == "STATE"
                                                                 ,UNIT_DESC == "ACRES"
                                                                 )

unique(data_usda_AG_LAND_AREA_2$DOMAIN_DESC)

data_usda_AG_LAND_AREA_2 <- data_usda_AG_LAND_AREA_2 %>% filter(
  DOMAIN_DESC == "AREA CROPLAND, HARVESTED" | DOMAIN_DESC == "AREA OPERATED AND AREA CROPLAND, HARVESTED"
)


#View(data_usda_AG_LAND_AREA_IOWA)

# only do numeric conversion here
data_usda_AG_LAND_AREA_2$VALUE <- as.numeric(gsub("[^0-9.]", "", data_usda_AG_LAND_AREA_2$VALUE))

testt <- data_usda_AG_LAND_AREA_2 %>%
  group_by(YEAR, STATE_NAME, DOMAIN_DESC) %>%
  summarize(land_area = sum(VALUE)) 

#View(testt)
options(repr.plot.width = 30, repr.plot.height = 30)

testt %>% ggplot(aes(x = YEAR, y = land_area)) + geom_bar(stat = "identity", aes(fill = DOMAIN_DESC), position = "dodge") + facet_wrap(~STATE_NAME, scales = "fixed") + theme_minimal() + theme(legend.position = "bottom", text = element_text(size = 5)) 


#data_usda_AG_LAND_AREA_IOWA %>% ggplot(aes(x = YEAR, y = VALUE, color = DOMAIN_DESC)) + geom_point()


```

```{r, message=FALSE, echo=FALSE}
data_usda_ENERGY <- data_usda %>% filter(GROUP_DESC == "ENERGY"
                                       ,COMMODITY_DESC == "ENERGY"
                                       ,STATISTICCAT_DESC == "EXPENSE" | STATISTICCAT_DESC == "SAVINGS"
                                       )

data_usda_ENERGY_IOWA <- data_usda_ENERGY %>% filter(STATE_NAME == "IOWA", AGG_LEVEL_DESC == "STATE"
                                                                 ,UNIT_DESC == "$ / OPERATION"
                                                                 )

# only do numeric conversion here
data_usda_ENERGY_IOWA$VALUE <- as.numeric(gsub("[^0-9.]", "", data_usda_ENERGY_IOWA$VALUE))

testt <- data_usda_ENERGY_IOWA %>%
  group_by(YEAR, STATISTICCAT_DESC) %>%
  summarize(dollars_per_operation = sum(VALUE)) 



testt %>% ggplot(aes(x = factor(YEAR), y = dollars_per_operation)) + geom_bar(stat = "identity", aes(fill = STATISTICCAT_DESC), position = "dodge") #+ coord_flip()

```


```{r, message=FALSE, echo=FALSE}
data_usda_ENERGY <- data_usda %>% filter(GROUP_DESC == "ENERGY"
                                       ,COMMODITY_DESC == "ENERGY"
                                       ,STATISTICCAT_DESC == "EXPENSE" | STATISTICCAT_DESC == "SAVINGS"
                                       )

data_usda_ENERGY_TEXAS <- data_usda_ENERGY %>% filter(STATE_NAME == "TEXAS", AGG_LEVEL_DESC == "STATE"
                                                                 ,UNIT_DESC == "$ / OPERATION"
                                                                 )

unique(data_usda_ENERGY_TEXAS$YEAR)




#View(data_usda_AG_LAND_AREA_IOWA)

# only do numeric conversion here
data_usda_ENERGY_TEXAS$VALUE <- as.numeric(gsub("[^0-9.]", "", data_usda_ENERGY_TEXAS$VALUE))

testt <- data_usda_ENERGY_TEXAS %>%
  group_by(YEAR, STATISTICCAT_DESC) %>%
  summarize(dollars_per_operation = sum(VALUE)) 

#View(testt)

testt %>% ggplot(aes(x = factor(YEAR), y = dollars_per_operation)) + geom_bar(stat = "identity", aes(fill = STATISTICCAT_DESC), position = "dodge") #+ coord_flip()


#data_usda_AG_LAND_AREA_IOWA %>% ggplot(aes(x = YEAR, y = VALUE, color = DOMAIN_DESC)) + geom_point()
```

```{r, message=FALSE, echo=FALSE}

data_usda_ENERGY <- data_usda %>% filter(GROUP_DESC == "ENERGY"
                                       ,COMMODITY_DESC == "ENERGY"
                                       ,STATISTICCAT_DESC == "EXPENSE" | STATISTICCAT_DESC == "SAVINGS"
                                       )

data_usda_ENERGY_ALL <- data_usda_ENERGY %>% filter(AGG_LEVEL_DESC == "STATE"
                                                                 ,UNIT_DESC == "$ / OPERATION"
                                                                 )

# only do numeric conversion here
data_usda_ENERGY_ALL$VALUE <- as.numeric(gsub("[^0-9.]", "", data_usda_ENERGY_ALL$VALUE))

testt <- data_usda_ENERGY_ALL %>%
  group_by(YEAR, STATE_NAME, STATISTICCAT_DESC) %>%
  summarize(dollars_per_operation = sum(VALUE)) 

options(repr.plot.width = 30, repr.plot.height = 30)

testt %>% ggplot(aes(x = YEAR, y = dollars_per_operation)) + geom_bar(stat = "identity", aes(fill = STATISTICCAT_DESC), position = "dodge") + facet_wrap(~STATE_NAME, scales = "fixed") + theme_minimal() + theme(legend.position = "bottom", text = element_text(size = 5))


```

```{r, message=FALSE, echo=FALSE}

data_usda_ENERGY <- data_usda %>% filter(GROUP_DESC == "ENERGY"
                                       ,COMMODITY_DESC == "ENERGY"
                                       ,STATISTICCAT_DESC == "EXPENSE" | STATISTICCAT_DESC == "SAVINGS"
                                       )

data_usda_ENERGY_ALL <- data_usda_ENERGY %>% filter(AGG_LEVEL_DESC == "STATE"
                                                                 ,UNIT_DESC == "$ / OPERATION"
                                                                 )

# only do numeric conversion here
data_usda_ENERGY_ALL$VALUE <- as.numeric(gsub("[^0-9.]", "", data_usda_ENERGY_ALL$VALUE))

data_usda_ENERGY_ALL_SAVINGS <- data_usda_ENERGY_ALL %>% filter(STATISTICCAT_DESC == "SAVINGS") 

data_usda_ENERGY_ALL_SAVINGS %>% group_by(STATE_NAME) %>% summarize(dollars_per_operation = sum(VALUE)) %>% 
  ggplot(aes(x = reorder(STATE_NAME,-dollars_per_operation), y = dollars_per_operation)) + geom_bar(stat = "identity") + coord_flip() + theme(text = element_text(size = 5))

data_usda_ENERGY_ALL_EXPENSES <- data_usda_ENERGY_ALL %>% filter(STATISTICCAT_DESC == "EXPENSE")

data_usda_ENERGY_ALL_EXPENSES %>% group_by(STATE_NAME) %>% summarize(dollars_per_operation = sum(VALUE)) %>% 
  ggplot(aes(x = reorder(STATE_NAME,-dollars_per_operation), y = dollars_per_operation)) + geom_bar(stat = "identity") + coord_flip() + theme(text = element_text(size = 5))

```

## MAZIN MOHAMMED

```{r, message=FALSE, echo=FALSE}
data_usda_preview$VALUE <- as.numeric(gsub("[^0-9.]", "", data_usda_preview$VALUE))


cattle_data <- data_usda_preview %>%
  filter(COMMODITY_DESC == "CATTLE") %>%
  select(COMMODITY_DESC, YEAR, STATE_NAME, VALUE, UNIT_DESC) %>%
  mutate(VALUE = as.numeric(VALUE)) 
average_cattle_value_by_state <- cattle_data %>%
  group_by(STATE_NAME) %>%
  summarise(Average_Value = mean(VALUE, na.rm = TRUE), .groups = 'drop') %>%
  arrange(desc(Average_Value))


print(head(average_cattle_value_by_state, n = 20))
```

```{r, message=FALSE, echo=FALSE}
# Assuming 'data_usda_preview' is already loaded and contains the necessary columns
# Filter for 'CATTLE' at the national level
national_cattle_data <- data_usda_preview %>%
  filter(COMMODITY_DESC == "CATTLE", STATE_NAME == "US TOTAL") %>%
  mutate(YEAR = as.numeric(YEAR),  # Ensure YEAR is numeric
         VALUE = as.numeric(gsub("[^0-9.]", "", VALUE)))  # Clean and convert VALUE to numeric

# Calculate the total production value by year for 'CATTLE' at the national level
cattle_trends_national <- national_cattle_data %>%
  group_by(YEAR) %>%
  summarise(Total_Value = sum(VALUE, na.rm = TRUE), .groups = 'drop') %>%
  arrange(YEAR)

# Plot the trend over years
ggplot(cattle_trends_national, aes(x = YEAR, y = Total_Value)) +
  geom_line() +
  geom_point() +  # Add points to visualize each year's data
  theme_minimal() +
  labs(title = "Trend of Cattle Production Value at the National Level",
       x = "Year",
       y = "Total Production Value ($)")
```


## Hazer Becic
Question: How does the distribution of agricultural expenses vary across different states in the United States over the years?

```{r, message=FALSE, echo=FALSE}
agri_expenses <- data_usda %>%
  filter(GROUP_DESC == "EXPENSES")

agg_expenses <- agri_expenses %>%
  group_by(STATE_NAME, YEAR) %>%
  summarise(total_expenses = sum(as.numeric(gsub(",", "", VALUE))))

# HAS WARNINGS  
ggplot(agg_expenses, aes(x = YEAR, y = total_expenses, color = STATE_NAME)) +
  geom_line() +
  labs(title = "Total Agricultural Expenses Over Years by State",
       x = "Year",
       y = "Total Expenses",
       color = "State") +
  theme_minimal() +
  theme(legend.position = "bottom")  # Adjusts the legend position to the bottom
```



Question: What are the main drivers of agricultural income variation across different regions in the United States, and how have these drivers evolved over time? 
```{r, message=FALSE, echo=FALSE}
agri_income <- data_usda %>%
  filter(GROUP_DESC == "INCOME")

# DOESN'T WORK
# Step 2: Explore the distribution of agricultural income across different regions and years
income_summary <- agri_income %>%
  group_by(REGION_DESC, YEAR) %>%
  summarise(total_income = sum(as.numeric(gsub(",", "", VALUE))))

# Step 3: Correlation analysis - Identify potential drivers of income variation
# For example, let's explore the correlation between income and commodity prices
# Assuming you have another dataset named commodity_prices containing commodity price data

# Merge commodity price data with income data based on common variables (e.g., year and commodity)
merged_data <- inner_join(income_summary, commodity_prices, by = c("YEAR", "COMMODITY_DESC"))

# Calculate correlation between income and commodity prices
correlation <- cor(merged_data$total_income, merged_data$price)

# Step 4: Data Visualization
# Plotting total agricultural income over the years for each region
ggplot(income_summary, aes(x = YEAR, y = total_income, color = REGION_DESC)) +
  geom_line() +
  labs(title = "Total Agricultural Income Over Years by Region",
       x = "Year",
       y = "Total Income",
       color = "Region") +
  theme_minimal()
```
